---
title: "Winter Habitat Use by Common Loons (Gavia Immer)"
subtitle: "on Mount Desert Island, Maine"
author: "Ellie Gabrielson, Will Draxler, Autumn Pauly"
date: "03/12/2024"
output: 
  html_document: 
    toc: true
---
# Background Information
## Introduction
Although Common Loons (Gavia immer) have been studied extensively during the breeding season not much is known of their behavior on their ocean wintering grounds, where Loons spend a significant part of their lives (Bent, 2009). The few studies that do look at winter behavior often yield conflicting results (McIntyre, 1978; Daub, 1989; Ford and Gieg, 1995). The possible use of different shoreline habitats for activities such as feeding, sleeping, or taking shelter is worthy of study. McIntyre (1978) found common loons were semi-social and formed loose rafts and that loon behavior varied with tides, weather, and time of day. Daub (1989) and later Ford and Gieg (1995) found refuting evidence suggesting loons were neither social nor territorial, and did not change their behavior much by tide. Holm and Burger (2002) found that the general pattern for loons was that they foraged during slack and main-flow water. Additionally, the possible difference in dive times as a result of varying tidal heights is worthy of study. Thompson and Price (2006) observed that dive times were longer during low tide in comparison to other tidal stages, though more studies should be conducted to assess the use of tidal stage and behavior. 

Our general hypothesis when analyzing this data is that environmental conditions affect the behavior of the *G. immer*. The null hypothesis to this statement is that environmental conditions do not affect the behavior of the *G. immer*.

## Methods
In 2023, observations were conducted at nine established observation sites along the coastline of Mount Desert Island in Hancock County (Maine). Observations occurred from January 2023 to March 2023 on a biweekly basis, where each waterfowl individual was recorded as well as their behaviors, referencing Daub’s (1989) list of behaviors, and coordinate location. Locational coordinates were obtained by estimating the individual’s rough location digitally (accurate to + or - 10 m). Each transect site was visited for roughly 5-10 minutes, where data collection followed initial observation. Utilizing a Kestrel 2500, weather data was recorded accurately to + or - 1°C of temperatures and within 3% of wind speed. Referencing Daub’s (1989) list of behaviors, each species’ behavior during the observation was noted. The weather and environmental data recorded included wind speed and direction, barometer, cloud cover, precipitation, humidity, temperature, tidal percentage, and wave class (Beaufort Wind Speed). Tidal data was gathered from NOAA’s digital tidal charts. 

In 2024, this study was repeated with the addition of observational variables including species dive times and tidal height. Dive times were recorded by using binoculars and a handheld timing watch to observe and time an individual several times. Tidal height was gathered from NOAA's digital tidal chart. 

# Data Analysis
## Load Packages and Data
```{r load-packages, message = FALSE, warning=FALSE}
#install.packages("tidyverse")
library(tidyverse)
library(rmarkdown)
library(scales)
library(ggridges)
library(rstatix)
library(parsnip)
```

```{r read-in-data, message = FALSE, warning=FALSE}
loons_2023 <- read_csv("loon_2023_tidy.csv")
```

### Diving Loons 2024

```{r loons-dive-2024}
# loons_dive_2024 <- loons_2024 %>% 
#   filter(species == "common_loon", 
#          behavior == "diving")
```

```{r}
# loons_dive_2024 <- loons_dive_2024 %>% 
#   separate_longer_delim(dive_time_obs, delim = ", ") %>% 
#   separate_longer_delim(dive_time_obs, delim = ",") %>% 
  
  
# loons_dive_2024 <- loons_dive_2024
#   mutate(dive_time_obs = as.numeric(dive_time_obs))

```


## General Statistics
The mean dive time of loons during this study was 41.95 seconds with a standard error of 1.53.

```{r}
#summary of dive times in 2024
summary(loons_dive_2024)

#standard error function
std.error <- function(x) sd(x)/sqrt(length(x))

std.error(loons_dive_2024$dive_time_obs)
```

There was an average of 9.55 loons observed per day during this study, with there being 2.12 being observed at each site.

```{r}
loons_2023 %>% 
  summary()

```













## General Tidal Stage and Behavior for 2023
The first analysis that will be conducted will be investigating the potential relationship between general tidal stages and *G. immer* behavior. As stated above, the behaviors chosen are referencing Daub’s (1989) list of behaviors. The tidal condition was sorted into three categories: `high`, `mid`, and `low`. 

It was hypothesized that tidal height would significantly affect the behavior of *G. immer* The null hypothesis stated that tidal height would not have a significant effect on *G. immer* behavior.

### Statistics
#### Chi-Squared Test of Independence
The Chi-Squared Test of Independence examines whether there is a relationship between two categorical variables. Our null hypothesis, that tidal height will not have a significant effect on *G. immer* behavior, assumes that tidal condition and behavior are independent. Our hypothesis, that tidal height will have a significant effect on *G. immer* behavior, assumes that tidal condition and behavior are dependent.

#### Expected Frequencies and Contingency Table
Before performing the chi-square test, we need to calculate the expected frequencies for each cell in the contingency table. These expected frequencies represent what you would expect to observe if there were no association between tidal condition and behavior.

```{r contingency-table, echo=FALSE}
loons_2023_behavior <- loons_2023 %>% select(behavior, general_tide)
loons_2024_behavior <- loons_dive_2024 %>% select(behavior, general_tide)
loons_behavior <- bind_rows(loons_2023_behavior, loons_2024_behavior)
contingency_table1 <- table(loons_behavior$general_tide, loons_behavior$behavior)
contingency_table1
```

#### Chi-square Test
Now that we have the contingency table of expected frequencies, we can perform the statistical test. 

```{r chi-square-test, warning=FALSE, echo=FALSE}
chi_sq_test1 <- chisq.test(contingency_table1)	
chi_sq_test1
```

REPORT: A Chi-squared test indicated a significant relationship between *G. immer* behavior and general tidal stage (X^2(12, n=307) = 75.8, df = 12, p-value > 0.005). This indicates that there is a relationship between general tidal stage and *G. immer* behavior, allowing the null hypothesis to be rejected.












## General Tidal Stage and Abundance at each site
The next analysis that will be conducted will be investigating the potential relationship between general tidal stages and *G. immer* abundance at each site.

The hypothesis is that tidal height would significantly affect the abundance of *G. immer* The null hypothesis stated that tidal height would not have a significant effect on *G. immer* abundance at each site.

### Statistics
#### Chi-Squared Test of Independence
The Chi-Squared Test of Independence examines whether there is a relationship between two categorical variables.

#### Expected Frequencies and Contingency Table
Before performing the chi-square test, we need to calculate the expected frequencies for each cell in the contingency table. These expected frequencies represent what you would expect to observe if there were no association between tidal condition and behavior.

```{r contingency-table, echo=FALSE}
contingency_table_abun <- table(loons_2023$tide, loons_2023$location)
contingency_table_abun

loons_2024_loon <- loons_2024 %>% 
  filter(species == "common_loon")

contingency_table_abun24 <- table(loons_2024_loon$tide, loons_2024_loon$location)
contingency_table_abun24
```

#### Chi-square Test
Now that we have the contingency table of expected frequencies, we can perform the statistical test. 

```{r chi-square-test, warning=FALSE, echo=FALSE}
chi_sq_testabun <- chisq.test(contingency_table_abun)
chi_sq_testabun

chisq24abun <- chisq.test(contingency_table_abun24)
chisq24abun
```

REPORT: A Chi-squared test indicated no significant relationship between *G. immer* abundance and tidal condition at each site (X^2(16, n=131) = 26.28, df = 16, p-value = 0.05017). This indicates that there is no relationship between tidal height and *G. immer* abundance at each site, allowing us to reject the hypothesis. 




















## Tidal Conditions and Dive Times in 2024
This study aimed to investigate the potential effect of tidal stage on the dive times of *G. immer*. The hypothesis posited that low tide would result in significantly longer dive times for *G. immer*. Conversely, the null hypothesis stated that tidal stage would not have a significant impact on *G. immer* dive times.


```{r message=FALSE, echo=FALSE, warning=FALSE}
library(stringr)
library(ggpubr)
library(tidymodels)
library(tidyverse)
```

```{r mutation, baby, echo=FALSE, message=FALSE, warning=FALSE}
loons_dive_2024 <- loons_2024 %>% 
  filter(species == "common_loon") %>% 
  select(tide, behavior, dive_time_obs, tidal_height, latitude, longitude) %>% 
  filter(dive_time_obs != "n/a") %>% 
  filter(dive_time_obs != "na")

loons_dive_2024 <- loons_dive_2024 %>% 
  separate_longer_delim(dive_time_obs, delim = ",")

loons_dive_2024 <- loons_dive_2024 %>% 
  separate_wider_delim(latitude, delim = ",", names = c("latitude", "longitude2"), too_few = "align_start") %>% 
  mutate(longitude2 = as.numeric(longitude2)) %>% 
  mutate(latitude = as.numeric(latitude))

loons_dive_2024 <- loons_dive_2024 %>%
  mutate(dive_time_obs = as.numeric(dive_time_obs))

loons_dive_2024 <- loons_dive_2024 %>% 
  mutate(general_tide = fct_recode(tide, "low" = "low_flood", 
                           "low" = "low_slack",
                           "mid" = "mid_flood",
                           "mid" = "mid_ebb",
                           "high" = "high_flood",
                           "high" = "high_slack",
                           "high" = "high_ebb"))
```


```{r}
#This is specifically to export coordinates, I know it's big but I didn't have a better solution

loons_dive_2024_1 <- loons_dive_2024 %>% 
  filter(is.na(longitude2) == F ) %>% 
  mutate(longitude = longitude2) %>% 
  select(-longitude2)

loons_dive_2024_2 <- loons_dive_2024 %>% 
  filter(is.na(longitude) == F ) %>% 
  select(- longitude2)

loons_dive_2024 <- loons_dive_2024_1 %>% 
  bind_rows(loons_dive_2024_2) %>% 
  write_csv("loons_dive_2024_GPS.csv")

```

### Visualizations
As can be seen below, we do not have a normal distribution for the variable `dive_time_obs`. We will need to perform a non-parametric test. 

```{r warning=TRUE, echo = FALSE}
ggplot(data = loons_dive_2024, mapping = aes(x = dive_time_obs)) +
  geom_histogram()+
  facet_wrap(~general_tide, ncol = 1) + 
  labs(title = "Distribution of Dive Times of G. Immer", 
       subtitle = "At various tidal stages", 
       x = "Dive Time (seconds)", 
       y = "Count")
```

#### Statistics 
The non-parametric nature of our dataset suggests that we need to use the Kruskal-Wallis H test (the "one-way ANOVA on ranks"), which is the non-parametric alternative to the one-way ANOVA test. 

```{r, warning=FALSE, echo=FALSE}
library(rstatix)
dive_kruskal <- kruskal.test(dive_time_obs ~ general_tide, data = loons_dive_2024)
dive_kruskal
```
```{r , warning=FALSE, echo=FALSE, message=FALSE}
#post-hoc test
dive_dunn <- loons_dive_2024 %>% dunn_test(dive_time_obs ~ general_tide, p.adjust.method = "bonferroni")
dive_dunn
```

```{r , echo=FALSE, warning=FALSE}
#This is not working - putting in code jail
# library(ggpubr)
# divetimeggline <- ggline(loons_dive_2024, x = "general_tide", y = "dive_time_obs",
#        add = c(add = "mean_se", "jitter"),
#        order = c("high", "mid", "low"),
#        ylab = "Tidal Stage", xlab = "Dive Time (seconds)", title = "")
# 
# divetimeggline + stat_compare_means(label.y = 75)
```

REPORT: The Kruskal-Wallis test revealed that there was no significant difference in dive times among the various tidal stages (Kruskal-Wallis rank sum test; H(2) = 5.82, P = 0.054). This suggests that tidal stage does not have a statistically significant effect on *G. immer* dive times, allowing us to reject our hypothesis. This implies that other factors besides tidal stage could be influencing dive behavior in *G. immer*. 

















## Site Location and Dive Times in 2024

This study aimed to investigate the potential effect of site locations on the dive times of *G. immer*. The hypothesis posited that sites would result in significantly different dive times for *G. immer*. Conversely, the null hypothesis stated that location would not have a significant impact on *G. immer* dive times.

```{r}
loons_2024_loon <- loons_2024_loon %>% 
  separate_longer_delim(dive_time_obs, delim = ",") %>% 
  mutate(dive_time_obs = as.numeric(dive_time_obs))
```


### Visualizations
As can be seen below, we do not have a normal distribution for the variable `dive_time_obs`. We will need to perform a non-parametric test. 

```{r warning=TRUE, echo = FALSE}
ggplot(data = loons_2024_loon, mapping = aes(x = dive_time_obs)) +
  geom_histogram()+
  labs(title = "Distribution of Dive Times of G. Immer", 
       subtitle = "At various tidal stages", 
       x = "Dive Time (seconds)", 
       y = "Count")
```

#### Statistics 
The non-parametric nature of our dataset suggests that we need to use the Kruskal-Wallis H test (the "one-way ANOVA on ranks"), which is the non-parametric alternative to the one-way ANOVA test. 

```{r, warning=FALSE, echo=FALSE}
library(rstatix)
site_kruskal <- kruskal.test(dive_time_obs ~ location, data = loons_2024_loon)
site_kruskal
```

```{r , warning=FALSE, echo=FALSE, message=FALSE}
#post-hoc test
site_dunn <- loons_2024_loon %>% dunn_test(dive_time_obs ~ location, p.adjust.method = "bonferroni")
site_dunn
```

```{r , echo=FALSE, warning=FALSE}
library(ggpubr)
divetimeggline <- ggline(loons_2024_loon, x = "location", y = "dive_time_obs",
       add = c(add = "mean_se", "jitter"),
       ylab = "Location", xlab = "Dive Time (seconds)", title = "")

divetimeggline + stat_compare_means(label.y = 75)

ggplot(data = loons_2024_loon, mapping = aes(x = location, y = dive_time_obs)) + 
  geom_boxplot()
```

REPORT: The Kruskal-Wallis test revealed that there was no significant difference in dive times among the various tidal stages (Kruskal-Wallis rank sum test; H(2) = 5.82, P = 0.054). This suggests that tidal stage does not have a statistically significant effect on *G. immer* dive times, allowing us to reject our hypothesis. This implies that other factors besides tidal stage could be influencing dive behavior in *G. immer*.








## Exposure Level and Abundance in 2023
This study aimed to explore the relationship between coastal exposure level and the presence of *G. immer*. The hypothesis suggested that higher exposure levels of coastal locations would positively affect the presence of *G. immer*. Conversely, the null hypothesis posited that coastal exposure level would not significantly influence *G. immer* presence.

### Visualizations
Below we are visualizing abundance in the form of boxplots. 

As we can see below, it appears as though there is the highest abundance at locations that are exposed and the lowest abundance at locations that are sheltered. Statistical tests will need to be performed to see if there is a significant difference in the abundances at different exposure levels. 

```{r warning=FALSE, echo=FALSE}
loons_2023 %>% 
  ggplot(mapping = aes(x = shelter_gradient, y = specific_abundance)) + 
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Common Loon Abundance", 
       subtitle = "by Shelter Gradient", 
       x = "Shelter Gradient", 
       y = "Abundance")

ggboxplot(loons_2023, x = "shelter_gradient", y = "specific_abundance", add = "jitter", xlab="Shelter Gradient", ylab="Abundance", order = c("exposed", "moderately_exposed", "moderate", "moderately_sheltered", "sheltered"))
```

```{r warning=FALSE, echo=FALSE}
loons_2023 %>% 
  ggplot(mapping = aes(x = specific_abundance)) +
  geom_histogram() +
  geom_density() + 
  facet_wrap(~shelter_gradient, ncol = 2)
```


```{r warning=FALSE, echo=FALSE}
#plot
loons_2023 %>% 
  ggplot(aes(x = specific_abundance, y = shelter_gradient, fill = shelter_gradient)) + 
  geom_density_ridges(alpha = 0.6, bins = 1) + 
  theme_ridges() + 
   theme(legend.position = "none", panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8))+
  theme_minimal() +
  scale_fill_viridis_d() +
  labs(title = "Common Loon Abundance", 
       subtitle = "by Shelter Gradient", 
       x = "Abundance", 
       y = "Shelter Gradient", 
       fill = "Shelter Gradient")
```

### Statistics
To determine if there is a significant difference in the abundance of *G. immer* in locations with varying exposure levels, where the exposure levels are categorical variable with five levels (exposed, moderately exposed, moderate, moderately sheltered, and sheltered), and the abundance is a continuous numerical variable, a one-way analysis of variance (ANOVA) will be used. 

```{r}
abundanceexposureaov = aov(specific_abundance ~ shelter_gradient, data = loons_2023)
summary(abundanceexposureaov)
TukeyHSD(abundanceexposureaov)
```

REPORT: The One-Way ANOVA revealed a significant effect of coastal exposure level on the abundance of *G. immer* (ANOVA, F_4,126 = 6.167; p = 0.000145). A Tukey HSD post-hoc comparison demonstrated that there was a significantly higher abundance of *G. immer* at exposed locations in comparison to moderately exposed locations. Additionally, there was a significantly higher abundance of *G. immer* at exposed locations in comparision to moderately sheltered locations. This indicates that coastal exposure level does indeed have a statistically significant influence on G. immer presence, allowing us to reject the null hypothesis. 







## Exposure Level and Behavior
This study aimed to investigate the potential influence of location exposure level on the behavior of Gavia immer (common loon). The hypothesis proposed that exposure level of the location significantly impacts G. immer behavior. In contrast, the null hypothesis stated that there would be no significant impact of location exposure level on G. immer behavior. 

### Visualizations

```{r}
loons_2023 %>% 
  count(behavior, shelter_gradient) %>% 
  ggplot(aes(x = behavior, y = n)) +
  geom_col(fill = "white", color = "black") +
  facet_wrap(~ shelter_gradient, nrow = 1) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Count of Common Loon Behavior", 
       subtitle = "by Shelter Gradient", 
       x = "Behavior", 
       y = "Count")
```

### Statistics
#### Chi-Squared Test of Independence
The Chi-Squared Test of Independence examines whether there is a relationship between two categorical variables. Our null hypothesis is that exposure will not have a significant effect on *G. immer* behavior assumes that exposure level and behavior are independent.

#### Expected Frequencies and Contingency Table
Before performing the chi-square test, we need to calculate the expected frequencies for each cell in the contingency table. These expected frequencies represent what you would expect to observe if there were no association between exposure and behavior.

```{r contingency-table2, echo=FALSE}
contingency_table2 <- table(loons_2023$shelter_gradient, loons_2023$behavior)
contingency_table2
```

#### Chi-square Test
Now that we have the contingency table of expected frequencies, we can perform the statistical test. 

```{r chi-square-test2, warning=FALSE, echo=FALSE}
chi_sq_test2 <- chisq.test(contingency_table2)
chi_sq_test2
```

REPORT: The Chi-Squared test revealed no significant association between exposure level of the location and *G. immer* behavior (X^2(16,n=131) = 16.77, p = 0.40). This suggests that exposure level of the location does not have a statistically significant impact on *G. immer* behavior, allowing us to reject the hypothesis. This implies that other factors may be more influential in determining behavioral patterns in *G. immer*. 

## Wave Class and Behavior
This study aimed to investigate the potential influence of wave class on the behavior of *G. immer*. The hypothesis proposed that wave class significantly affects *G. immer* behavior. Conversely, the null hypothesis stated that there would be no significant effect of wave class on *G. immer* behavior.

### Visualizations

```{r}
loons_2023 %>% 
 ggplot(aes(x = wave_class)) +
  geom_histogram() +
  facet_wrap(~ behavior, nrow = 1)

loons_2023 %>%
  ggplot(aes(x = wave_class, y = behavior, fill = behavior)) + 
  geom_density_ridges(alpha = 0.6, bins = 1)
```

### Statistics

```{r}
waveclassaov = aov(wave_class ~ behavior, data = loons_2023)
summary(waveclassaov)
```


REPORT: Wave class did not significantly affect *G. immer* behavior (one-way ANOVA, F_4,126 = 2.356, P = 0.0572).

### Statistics
#### Chi-Squared Test of Independence
The Chi-Squared Test of Independence examines whether there is a relationship between two categorical variables. 

#### Expected Frequencies and Contingency what is Table
Before performing the chi-square test, we need to calculate the expected frequencies for each cell in the contingency table. These expected frequencies represent what you would expect to observe if there were no association between wave class and behavior.

```{r contingency-table2, echo=FALSE}
contingency_table3 <- table(loons_2023$wave_class, loons_2023$behavior)
contingency_table3
```

#### Chi-square Test
Now that we have the contingency table of expected frequencies, we can perform the statistical test. 

```{r chi-square-test2, warning=FALSE, echo=FALSE}
chi_sq_test3 <- chisq.test(contingency_table3)
chi_sq_test3
```

REPORT: The Chi-Squared test revealed that wave class did not significantly affect *G. immer* behavior in our sample (Chi-squared(24, n=131) = 28.64, p = 0.23). This suggests that there is no statistically significant relationship between wave class and *G. immer* behavior, allowing us to reject the hypothesis. It is possible that other factors not accounted for in this study may have a greater influence on *G. immer* behavior. 

## Meters Offshore and Dive Times
This study aimed to explore the potential effect of distance from shore on the dive times of *G. immer*. The hypothesis suggested that distance from shore would significantly influence *G. immer* dive times. Conversely, the null hypothesis posited that there would be no significant effect of distance from shore on *G. immer* dive times. 

### Visualizations

```{r}
loons_dive_2024 <- loons_2024 %>% 
  filter(species == "common_loon") %>% 
  select(tide, behavior, dive_time_obs, tidal_height, meters_offshore) %>% 
  filter(dive_time_obs != "n/a") %>% 
  filter(dive_time_obs != "na")

loons_dive_2024 <- loons_dive_2024 %>% 
  separate_longer_delim(dive_time_obs, delim = ",") %>% 
  mutate(dive_time_obs = as.numeric(dive_time_obs))

loons_dive_2024 %>% 
  ggplot(aes(x = meters_offshore, y = dive_time_obs)) + 
  geom_jitter() + 
  geom_smooth(method = "lm")

#lm
lm3=lm(loons_dive_2024$meters_offshore~loons_dive_2024$dive_time_obs)
lm3

plot(loons_dive_2024$meters_offshore,loons_dive_2024$dive_time_obs,pch=16)
lm3=lm(loons_dive_2024$meters_offshore~loons_dive_2024$dive_time_obs)
abline(lm3)


#scatterplot
meters_offshore <- loons_dive_2024$meters_offshore
dive_time_obs <- loons_dive_2024$dive_time_obs

plot(meters_offshore, dive_time_obs, main="Scatter plot", xlab="Meters Offshore", ylab="Dive Time (second)", pch=19)
abline(lm(dive_time_obs~meters_offshore), col="black")
```

### Statistics
With both of our variables being continuous and numerical, we need to perform a regression test. First, we need to determine if there is a linear relationship between the variables. 

```{r}
loons_dive_2024 %>% 
  ggplot(aes(x = meters_offshore, y = dive_time_obs)) + 
  geom_jitter() + 
  geom_smooth(method = "lm") + 
  labs(title = "Dive Times vs. Meters Offshore", 
       subtitle = "of G. immer", 
       x = "Distance Offshore (m)", 
       y = "Dive Time (seconds)")
```

Next, we need to determine if both variables are normally distributed. 

```{r}
#Meters Offshore
ggplot(data = loons_dive_2024, mapping = aes(x = meters_offshore)) + 
  geom_histogram()

#Dive Times
ggplot(data = loons_dive_2024, mapping = aes(x = dive_time_obs)) + 
  geom_histogram()
```

#### Statistics
With the variable `meters_offshore` having a non-normal distribution, we will want to use a non-parametric version of a linear correlation test (i.e. Pearson's R). One such test is Spearman's Rho. 

```{r warning=FALSE}
dive_wave_corr <- cor.test(x=loons_dive_2024$meters_offshore, y = loons_dive_2024$dive_time_obs, method = 'spearman')

dive_wave_corr
```

REPORT: A Spearman's correlation revealed a moderate positive monotonic relationship between the variables `meters_offshore` and `dive_time_obs` (rs[93] = .50, p < 0.005). This indicates that there is a statistically significant association between distance from shore and G. immer dive times, allowing us to reject the null hypothesis. This implies that G. immer may exhibit longer dive times when further from the shore.

## Tidal height and dive times
This study aimed to investigate the potential effect of tidal height on the dive times of *G. immer*. The hypothesis proposed that tidal height would significantly influence *G. immer* dive times. Conversely, the null hypothesis stated that there would be no significant effect of tidal height on *G. immer* dive times.

### Visualizations
```{r}
loons_dive_2024 %>% 
  ggplot(aes(x = tidal_height, y = dive_time_obs)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

### Statistics
First, we need to see if our variables have a linear relationship

```{r}
loons_dive_2024 %>% 
  ggplot(aes(x = tidal_height, y = dive_time_obs)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

Next, we need to determine if both variables are normally distributed. 

```{r}
#tidal height
ggplot(data = loons_dive_2024, mapping = aes(x = tidal_height)) + 
  geom_histogram()

#dive times
ggplot(data = loons_dive_2024, mapping = aes(x = dive_time_obs)) + 
  geom_histogram()
```

#### Statistics
With the variable `tidal_height` having a non-normal distribution, we will want to use a non-parametric version of a linear correlation test, such as the Spearman's correlation. 

```{r warning=FALSE}
dive_height_corr <- cor.test(x=loons_dive_2024$tidal_height, y = loons_dive_2024$dive_time_obs, method = 'spearman')

dive_height_corr
```

REPORT: Spearman's correlation analysis indicated that there is no significant relationship between tidal height and *G. immer* dive times  (rs[93] = -0.06, p = 0.54). This suggests that tidal height does not have a statistically significant impact on *G. immer* dive times, allowing us to reject the hypothesis. This implies that other factors may have a greater influence on dive behavior in *G. immer*.



## Tidal classification and dive times
This study aimed to explore the potential effect of tidal class on the dive times of *G. immer*. The hypothesis posited that tidal class would significantly influence *G. immer* dive times. Conversely, the null hypothesis stated that there would be no significant effect of tidal class on *G. immer* dive times.

### Visualizations

```{r}
loons_dive_2024$tide <- fct_relevel(loons_dive_2024$tide, c("mid_ebb",
                                                            "low_slack",
                                                            "low_flood",
                                                            "mid_flood",
                                                            "high_flood",
                                                            "high_slack",
                                                            "high_ebb"))
loons_dive_2024 %>% 
ggplot(mapping = aes(x = tide, y = dive_time_obs)) + 
  geom_boxplot() + 
  theme_minimal() + 
  labs(title = "",
       x = "Specific Tidal Stage", 
       y = "Dive Time (seconds)")

library(ggpubr)
divetimeagainggline <- ggline(loons_dive_2024, x = "tide", y = "dive_time_obs",
       add = c(add = "mean_se", "jitter"),
       order = c("mid_ebb", "low_slack", "low_flood", "mid_flood", "high_flood", "high_slack", "high_ebb"),
       ylab = "Tidal Stage", xlab = "Dive Time (seconds)", title = "")
divetimeagainggline
```

### Statistics

Below are the statistics tests that test against all tidal stages. 
```{r}
full_tide_aov = aov(dive_time_obs ~ tide, data = loons_dive_2024)
summary(full_tide_aov)
TukeyHSD(full_tide_aov)
```

REPORT: The ANOVA test demonstrated that tidal class has a significant effect on the length of G. immer dive times (ANOVA, F_6,86 = 4.599; p = 0.000427). An unplanned analytical comparison demonstrated that *G. immer* dive times at low slack tide were significantly different than those at mid ebb tide (TukeyB, p < 0.005), and dive lengths at high ebb tide were significantly different than those at mid ebb tide (TukeyB, p < 0.05). Otherwise, there were no significant differences in dive times between other tidal classes (TukeyB, p > 0.05). These findings allow us to reject the null hypothesis. 

Below are the statistics tests that test against all tidal stages. 
```{r}
loons_dive_2024_2 <- loons_dive_2024 %>% 
  filter(tide == "low_slack"|tide == "high_flood"|tide == "low_flood") %>% 
  mutate(flood = if_else(tide == "high_flood"|tide == "low_flood", "flood", "low_slack"))

loons_dive_2024_2

t.test(dive_time_obs ~ flood, data = loons_dive_2024_2)
```

REPORT: A Welch's Two Sample t-test demonstrated that there was a significant difference in dive times during low-slack periods and flooding periods (df = 12.743, p = 0.04036). 

# Discussion / Conclusion
This study provides insights into the winter ecology and behavioral dynamics of the Common Loon (Gavia immer) along the coastline of Mount Desert Island, Maine. While much of the existing research has focused on the breeding season ecology of this species, this study sheds light on their non-breeding season behaviors, particularly in coastal wintering habitats.

Our findings reveal relationships between environmental factors and loon behavior, highlighting the multifaceted nature of ecological interactions shaping the dynamics of G. immer populations. Furthermore, our study highlights the significant role of habitat characteristics, particularly coastal exposure level, in shaping the spatial distribution and abundance of loon populations along the coastline. The observed preference of loons for exposed coastal environments underscores the importance of habitat structure and resource availability in mediating population dynamics and ecological interactions.

Further studies are warranted to determine more comprehensive ecological dynamics of loon populations in diverse coastal environments to ensure the long-term conservation and sustainability of this species.

# Citations
Bent, A. C. 2009. Life Histories of North American Diving Birds. Cornell University Press: Ithaca, New York.

Daub, B. C. 1989. Behavior of Common Loons in Winter. Journal of Field Ornithology 60: 305-311.

Ford, T. B. and Gieg, J. A. 1995. Winter Behavior of the Common Loon. Journal of Field Ornithology 66: 22-29.

McIntyre, J. W. 1978. Wintering Behavior of Common Loons. The Auk 95: 396-403.

Holm, K. and Burger, A. 2002. Foraging Behavior and Resource Partitioning by Diving Birds During Winter in Areas of Strong Tidal Currents. Waterbirds 25(3): 312-325.

Thompson, Stephanie A. & Price, J. J. (2006). Water Clarity and Diving Behavior in Wintering Common Loons. Waterbirds: The International Journal of Waterbird Biology, 29(2), 169–175.




### Randomly Sampling Data

```{r , warning=FALSE, echo=FALSE}
 # ggplot(data = loons_dive_2024, mapping = aes(x = dive_time_obs)) +
 #   geom_histogram()+
 #   facet_wrap(~general_tide, ncol = 1)
 # 
 # loons_dive_2024_means <- loons_dive_2024 %>%
 #   group_by(general_tide) %>%
 #   summarise(mean = mean(dive_time_obs))
```

```{r, warning=FALSE}
# library(infer)
# 
# #low tide
# loons_lowdive_2024 <- loons_dive_2024 %>%
#   filter(general_tide == "low") %>%
#   select(dive_time_obs)
# 
# set.seed(101)
# 
# sampled_low <- loons_lowdive_2024 %>%
#   specify(response = dive_time_obs) %>%
#   generate(reps = 1000, type = "bootstrap") %>%
#   calculate(stat = "mean")
# 
# sampled_low <- sampled_low %>%
#   mutate(replicate = as.character(replicate)) %>%
#   mutate(tide = if_else(replicate != 0, "low", replicate)) %>%
#   select(tide, stat)
# 
# 
# 
# #mid tide
# loons_middive_2024 <- loons_dive_2024 %>%
#   filter(general_tide == "mid") %>%
#   select(dive_time_obs)
# 
# set.seed(101)
# 
# sampled_mid <- loons_middive_2024 %>%
#   specify(response = dive_time_obs) %>%
#   generate(reps = 1000, type = "bootstrap") %>%
#   calculate(stat = "mean")
# 
# sampled_mid <- sampled_mid %>%
#   mutate(replicate = as.character(replicate)) %>%
#   mutate(tide = if_else(replicate != 0, "mid", replicate)) %>%
#   select(tide, stat)
# 
# 
# 
# #high tide
# loons_highdive_2024 <- loons_dive_2024 %>%
#   filter(general_tide == "high") %>%
#   select(dive_time_obs)
# 
# set.seed(101)
# 
# sampled_high <- loons_highdive_2024 %>%
#   specify(response = dive_time_obs) %>%
#   generate(reps = 1000, type = "bootstrap") %>%
#   calculate(stat = "mean")
# 
# sampled_high <- sampled_high %>%
#   mutate(replicate = as.character(replicate)) %>%
#   mutate(tide = if_else(replicate != 0, "high", replicate)) %>%
#   select(tide, stat)
# 
# #joining dataset
# dive_samples <- rbind(sampled_low, sampled_high)
# dive_samples <- rbind(dive_samples, sampled_mid)
```

Now, we can see that we have a normal distribution of dive times during all tidal cycles, which now makes an ANOVA an appropriate statistical test to use for this data frame. 

```{r sampledivetimes, warning=FALSE, echo=FALSE}
# ggplot(data = dive_samples, mapping = aes(x = stat)) +
#   geom_histogram() +
#   facet_wrap(~tide, ncol = 1) +
#   theme_minimal() +
#   labs(title = "Dive Times of Common Loons",
#        subtitle = "during low, mid, and high tide",
#        x = "Dive Time (seconds)",
#        y = "Count") +
#   geom_vline(data = loons_dive_2024_means, aes(xintercept = mean, color = general_tide))
```

```{r echo=FALSE, warning=FALSE}
# dive_samples$tide <- fct_relevel(dive_samples$tide, c("low", "mid", "high"))
# 
# ggplot(data = dive_samples, mapping = aes(x = tide, y = stat)) +
#   geom_boxplot() +
#   labs(title = "Tidal Conditions and Dive Times",
#        subtitle = "of Gavia immer",
#        x = "Tidal Stage",
#        y = "Dive Time (seconds)")
```

To determine if there is a significant difference in dive times of Common loons during different tidal stages, where the tidal stages are categorical with three levels (high tide, mid tide, and low tide), and the dive time is a continuous numerical variable, a one-way analysis of variance (ANOVA) will be used. 

```{r diveaov, echo=FALSE}
# diveaov = aov(stat ~ tide, data = dive_samples)
```

```{r anova2,warning=FALSE, echo=FALSE}
# summary(diveaov)
# TukeyHSD(diveaov)
```

```{r ,message=FALSE, echo=FALSE, warning=FALSE}
# #high to low
# loons_dive_2024_1 <- dive_samples %>%
#   filter(tide == "high"|tide == "mid")
# 
# t.test(stat ~ tide, data = loons_dive_2024_1)
# 
# #low to high
# loons_dive_2024_1 <- dive_samples %>%
#   filter(tide == "high"|tide == "low")
# 
# t.test(stat ~ tide, data = loons_dive_2024_1)
# 
# #low to mid
# loons_dive_2024_1 <- dive_samples %>%
#   filter(tide == "mid"|tide == "low")
# 
# t.test(stat ~ tide, data = loons_dive_2024_1)
```




```{r}
library(tidyverse)
dives_gis_tidy <- read_csv("dives_gis_tidy.csv") 


dives_gis_tidy_high <- dives_gis_tidy %>%
  filter(general_tide == "high") 

dives_gis_tidy_high %>% 
  ggplot(aes(x = water_depth_m, y = dive_time_obs)) + 
  geom_jitter() +
  geom_smooth(method = "lm") +
  labs(
    title = "High"
  )

dives_gis_tidy_high_lm <- lm(water_depth_m ~ dive_time_obs, data = dives_gis_tidy_high)
summary(dives_gis_tidy_high_lm)


dives_gis_tidy_mid <- dives_gis_tidy %>% 
  filter(general_tide == "mid")

dives_gis_tidy_mid %>% 
  ggplot(aes(x = water_depth_m, y = dive_time_obs)) + 
  geom_jitter() +
  geom_smooth(method = "lm") +
  labs(
    title = "Mid"
  )

dives_gis_tidy_mid_lm <- lm(water_depth_m ~ dive_time_obs, data = dives_gis_tidy_mid)
summary(dives_gis_tidy_mid_lm)

dives_gis_tidy_low <- dives_gis_tidy %>% 
  filter(general_tide == "low")

dives_gis_tidy_low %>% 
  ggplot(aes(x = water_depth_m, y = dive_time_obs)) + 
  geom_jitter() +
  geom_smooth(method = "lm") +
  labs(
    title = "Low"
  )

dives_gis_tidy_low_lm <- lm(water_depth_m ~ dive_time_obs, data = dives_gis_tidy_low)
summary(dives_gis_tidy_low_lm)


```































































