---
title: "Winter Habitat Use by Common Loons (Gavia Immer)"
subtitle: "on Mount Desert Island, Maine"
author: "Ellie Gabrielson, Will Draxler, Autumn Pauly"
date: "03/12/2024"
output: 
  html_document: 
    toc: true
---
# Background Information
## Introduction
Although Common Loons (Gavia immer) have been studied extensively during the breeding season not much is known of their behavior on their ocean wintering grounds, where Loons spend a significant part of their lives (Bent, 2009). The few studies that do look at winter behavior often yield conflicting results (McIntyre, 1978; Daub, 1989; Ford and Gieg, 1995). The possible use of different shoreline habitats for activities such as feeding, sleeping, or taking shelter is worthy of study. McIntyre (1978) found common loons were semi-social and formed loose rafts and that loon behavior varied with tides, weather, and time of day. Daub (1989) and later Ford and Gieg (1995) found refuting evidence suggesting loons were neither social nor territorial, and did not change their behavior much by tide. Holm and Burger (2002) found that the general pattern for loons was that they foraged during slack and main-flow water. Additionally, the possible difference in dive times as a result of varying tidal heights is worthy of study. Thompson and Price (2006) observed that dive times were longer during low tide in comparison to other tidal stages, though more studies should be conducted to assess the use of tidal stage and behavior. 

Our general hypothesis when analyzing this data is that environmental conditions affect the behavior of the *G. immer*. The null hypothesis to this statement is that environmental conditions do not affect the behavior of the *G. immer*.

## Methods
In 2023, observations were conducted at nine established observation sites along the coastline of Mount Desert Island in Hancock County (Maine). Observations occurred from January 2023 to March 2023 on a biweekly basis, where each waterfowl individual was recorded as well as their behaviors, referencing Daub’s (1989) list of behaviors, and coordinate location. Locational coordinates were obtained by estimating the individual’s rough location digitally (accurate to + or - 10 m). Each transect site was visited for roughly 5-10 minutes, where data collection followed initial observation. Utilizing a Kestrel 2500, weather data was recorded accurately to + or - 1°C of temperatures and within 3% of wind speed. Referencing Daub’s (1989) list of behaviors, each species’ behavior during the observation was noted. The weather and environmental data recorded included wind speed and direction, barometer, cloud cover, precipitation, humidity, temperature, tidal percentage, and wave class (Beaufort Wind Speed). Tidal data was gathered from NOAA’s digital tidal charts. This study was repeated in 2024 with the addition of observational variables including species dive times and tidal height. 

# Data Analysis
## Load Packages and Data
```{r load-packages, message = FALSE, warning=FALSE}
#install.packages("tidyverse")
library(tidyverse)
library(rmarkdown)
library(scales)
library(ggridges)
library(rstatix)
library(parsnip)
```

```{r read-in-data, message = FALSE, warning=FALSE}
loons_2023 <- read_csv("loon_2023_tidy.csv")
loons_2024 <- read_csv("loon_2024_tidy.csv")
```

## General Tidal Stage and Behavior for 2023

The first analysis that will be conducted will be investigating the potential relationship between general tidal stages and *G. immer* behavior. As stated above, the behaviors chosen are referencing Daub’s (1989) list of behaviors. The tidal condition was sorted into three categories: `high`, `mid`, and `low`. 

Hypothesis: General tidal stages have a significant effect on *G. immer* behavior. 
Null Hypothesis: General tidal stages will not have a significant effect on *G. immer* behavior.


### Statistics
#### Chi-Squared Test of Independence
The Chi-Squared Test of Independence examines whether there is a relationship between two categorical variables. Our null hypothesis, that tidal height will not have a significant effect on *G. immer* behavior, assumes that tidal condition and behavior are independent. Our hypothesis, that tidal height will have a significant effect on *G. immer* behavior, assumes that tidal condition and behavior are dependent.

#### Expected Frequencies and Contingency Table
Before performing the chi-square test, we need to calculate the expected frequencies for each cell in the contingency table. These expected frequencies represent what you would expect to observe if there were no association between tidal condition and behavior.

```{r contingency-table, echo=FALSE}
contingency_table1 <- table(loons_2023$tide, loons_2023$behavior)
contingency_table1
```

#### Chi-square Test
Now that we have the contingency table of expected frequencies, we can perform the statistical test. 

```{r chi-square-test, warning=FALSE, echo=FALSE}
chi_sq_test1 <- chisq.test(contingency_table1)
chi_sq_test1

# # x-squared
# chi_sq_test1$statistic
# 
# #p-value
# chi_sq_test1$p.value
# 
# #expected frequencies
# chi_sq_test1$expected
```

REPORT: The behavior of Common Loons is dependent on tidal condition (Pearson's Chi-squared test; X-squared = 33.184, df = 10, p-value > 0.005). 

## Tidal Conditions and Dive Times in 2024

The second analysis that will be conducted will investigate the potential relationship between general tidal stages and *G. immer* dive times. 

Hypothesis: Low tide will cause *G. immer* to have significantly longer dive times. 
Null Hypothesis: Low tide will not cause *G. immer* to have significantly longer dive times. 

```{r message=FALSE, echo=FALSE, warning=FALSE}
library(stringr)
library(ggpubr)
library(tidymodels)
```

```{r mutation, baby, echo=FALSE, message=FALSE, warning=FALSE}
loons_dive_2024 <- loons_2024 %>% 
  filter(species == "common_loon") %>% 
  select(tide, behavior, dive_time_obs, tidal_height) %>% 
  filter(dive_time_obs != "n/a") %>% 
  filter(dive_time_obs != "na")

loons_dive_2024 <- loons_dive_2024 %>% 
  separate_longer_delim(dive_time_obs, delim = ",")

loons_dive_2024 <- loons_dive_2024 %>%
  mutate(dive_time_obs = as.numeric(dive_time_obs))

loons_dive_2024 <- loons_dive_2024 %>% 
  mutate(general_tide = fct_recode(tide, "low" = "low_flood", 
                           "low" = "low_slack",
                           "mid" = "mid_flood",
                           "mid" = "mid_ebb",
                           "high" = "high_flood",
                           "high" = "high_slack",
                           "high" = "high_ebb"))
```

### Visualizations

As can be seen below, we do not have a normal distribution for the variable `dive_time_obs`. We will need to perform a non-parametric test. 

```{r warning=TRUE, echo = FALSE}
ggplot(data = loons_dive_2024, mapping = aes(x = dive_time_obs)) +
  geom_histogram()+
  facet_wrap(~general_tide, ncol = 1) + 
  labs(title = "Distribution of Dive Times of G. Immer", 
       subtitle = "At various tidal stages", 
       x = "Dive Time (seconds)", 
       y = "Count")
```

#### Statistics 

The non-parametric nature of our dataset suggests that we need to use the Kruskal-Wallis H test (the "one-way ANOVA on ranks"), which is the non-parametric alternative to the one-way ANOVA test. 

```{r, warning=FALSE, echo=FALSE}
library(rstatix)
dive_kruskal <- kruskal.test(dive_time_obs ~ general_tide, data = loons_dive_2024)
dive_kruskal
```
```{r , warning=FALSE, echo=FALSE, message=FALSE}
#post-hoc test
dive_dunn <- loons_dive_2024 %>% dunn_test(dive_time_obs ~ general_tide, p.adjust.method = "bonferroni")
dive_dunn
```

```{r , echo=FALSE, warning=FALSE}
library(ggpubr)
divetimeggline <- ggline(loons_dive_2024, x = "general_tide", y = "dive_time_obs",
       add = c(add = "mean_se", "jitter"),
       order = c("high", "mid", "low"),
       ylab = "Tidal Stage", xlab = "Dive Time (seconds)", title = "G. Immer Dive Times at Various Tidal Stages")

divetimeggline + stat_compare_means(label.y = 75)
```

REPORT: The dive times of *G. immer* does not differ significantly from various tidal stages (Kruskal-Wallis rank sum test; H(2) = 5.82, P = 0.054). 










## Exposure Level and Abundance in 2023

The third analysis that will be conducted will be investigating the potential relationship between exposure level and *G. immer* abundance during the 2023 observation period. The null hypothesis for this analysis suggests that *G. immer* abundance will be independent of exposure level, whilst the hypothesis suggests the opposite. 

Null Hypothesis: Higher exposure level of location will not significantly positively affect *G. immer* presence.  

Hypothesis: Higher exposure level of location significantly positively affects *G. immer* presence. 

### Visualizations
Below we are visualizing abundance in the form of boxplots. 

As we can see below, it appears as though there is the highest abundance at locations that are exposed and the lowest abundance at locations that are sheltered. Statistical tests will need to be performed to see if there is a significant difference in the abundances at different exposure levels. 

```{r warning=FALSE, echo=FALSE}
loons_2023 %>% 
  ggplot(mapping = aes(x = shelter_gradient, y = specific_abundance)) + 
  geom_boxplot() +
  labs(title = "Common Loon Abundance", 
       subtitle = "by Shelter Gradient", 
       x = "Shelter Gradient", 
       y = "Abundance")
```

```{r warning=FALSE, echo=FALSE}
loons_2023 %>% 
  ggplot(mapping = aes(x = specific_abundance)) +
  geom_histogram() +
  geom_density() + 
  facet_wrap(~shelter_gradient, ncol = 2)
```


```{r warning=FALSE, echo=FALSE}
#plot
loons_2023 %>% 
  ggplot(aes(x = specific_abundance, y = shelter_gradient, fill = shelter_gradient)) + 
  geom_density_ridges(alpha = 0.6, bins = 1) + 
  theme_ridges() + 
   theme(legend.position = "none", panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8))+
  theme_minimal() +
  scale_fill_viridis_d() +
  labs(title = "Common Loon Abundance", 
       subtitle = "by Shelter Gradient", 
       x = "Abundance", 
       y = "Shelter Gradient", 
       fill = "Shelter Gradient")
```

### Statistics
To determine if there is a significant difference in the abundance of *G. immer* in locations with varying exposure levels, where the exposure levels are categorical variable with five levels (exposed, moderately exposed, moderate, moderately sheltered, and sheltered), and the abundance is a continuous numerical variable, a one-way analysis of variance (ANOVA) will be used. 

```{r}
abundanceexposureaov = aov(specific_abundance ~ shelter_gradient, data = loons_2023)
summary(abundanceexposureaov)
TukeyHSD(abundanceexposureaov)
```

REPORT: A One-Way ANOVA demonstrated that coastal exposure level has a significant effect on the abundance of *G. immer* (ANOVA, F_4,126 = 6.167; p = 0.000145). An unplanned analytical comparison demonstrated that *G. immer* presence [finish this].

## Exposure Level and Behavior

The fourth analysis that will be conducted will be investigating the potential relationship between exposure level and *G. immer* behavior during the 2023 observation period. 

Null Hypothesis: Exposure level of location has no significant impact on *G. immer* behavior.
Hypothesis: Exposure level of location significantly impacts *G. immer* behavior. 

### Visualizations

```{r}
loons_2023 %>% 
  count(behavior, shelter_gradient) %>% 
  ggplot(aes(x = behavior, y = n)) +
  geom_col(fill = "white", color = "black") +
  facet_wrap(~ shelter_gradient, nrow = 1) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Count of Common Loon Behavior", 
       subtitle = "by Shelter Gradient", 
       x = "Behavior", 
       y = "Count")
```

### Statistics
#### Chi-Squared Test of Independence
The Chi-Squared Test of Independence examines whether there is a relationship between two categorical variables. Our null hypothesis is that exposure will not have a significant effect on *G. immer* behavior assumes that exposure level and behavior are independent.

#### Expected Frequencies and Contingency Table
Before performing the chi-square test, we need to calculate the expected frequencies for each cell in the contingency table. These expected frequencies represent what you would expect to observe if there were no association between exposure and behavior.

```{r contingency-table2, echo=FALSE}
contingency_table2 <- table(loons_2023$shelter_gradient, loons_2023$behavior)
contingency_table2
```

#### Chi-square Test
Now that we have the contingency table of expected frequencies, we can perform the statistical test. 

```{r chi-square-test2, warning=FALSE, echo=FALSE}
chi_sq_test2 <- chisq.test(contingency_table2)
chi_sq_test2
```

REPORT: 
A Pearson's Chi-squared test demonstrated that there was no significant association between exposure level and behavior (X^2(16,n=131) = 16.77, p = 0.40). 

## Wave Class and Behavior
The fifth analysis that will be conducted will be investigating the potential relationship between wave class and *G. immer* behavior during the 2023 observation period. 

Hypothesis: Wave class will significantly affect *G. immer* behavior.

Null Hypothesis: Wave class will not significantly affect *G. immer* behavior.

### Visualizations

```{r}
loons_2023 %>% 
 ggplot(aes(x = wave_class)) +
  geom_histogram() +
  facet_wrap(~ behavior, nrow = 1)

loons_2023 %>%
  ggplot(aes(x = wave_class, y = behavior, fill = behavior)) + 
  geom_density_ridges(alpha = 0.6, bins = 1)
```

### Statistics

```{r}
waveclassaov = aov(wave_class ~ behavior, data = loons_2023)
summary(waveclassaov)
```


REPORT: Wave class did not significantly affect *G. immer* behavior (one-way ANOVA, F_4,126 = 2.356, P = 0.0572).

### Statistics
#### Chi-Squared Test of Independence
The Chi-Squared Test of Independence examines whether there is a relationship between two categorical variables. 

#### Expected Frequencies and Contingency what is Table
Before performing the chi-square test, we need to calculate the expected frequencies for each cell in the contingency table. These expected frequencies represent what you would expect to observe if there were no association between wave class and behavior.

```{r contingency-table2, echo=FALSE}
contingency_table3 <- table(loons_2023$wave_class, loons_2023$behavior)
contingency_table3
```

#### Chi-square Test
Now that we have the contingency table of expected frequencies, we can perform the statistical test. 

```{r chi-square-test2, warning=FALSE, echo=FALSE}
chi_sq_test3 <- chisq.test(contingency_table3)
chi_sq_test3
```

REPORT: A chi-squared test demonstrated that wave class did not significantly affect *G. immer* behavior in our sample (Chi-squared(24, n=131) = 28.64, p = 0.23).








## Meters Offshore and Dive Times
The sixth analysis that will be conducted will be investigating the potential relationship between distance from shore and *G. immer* dive times during the 2024 observation period. 

Hypothesis: Distance from shore will significantly affect *G. immer* dive times. 
Null Hypothesis: Distance from shore will not significantly affect *G. immer* dive times.

### Visualizations

```{r}
loons_dive_2024 <- loons_2024 %>% 
  filter(species == "common_loon") %>% 
  select(tide, behavior, dive_time_obs, tidal_height, meters_offshore) %>% 
  filter(dive_time_obs != "n/a") %>% 
  filter(dive_time_obs != "na")

loons_dive_2024 <- loons_dive_2024 %>% 
  separate_longer_delim(dive_time_obs, delim = ",") %>% 
  mutate(dive_time_obs = as.numeric(dive_time_obs))

loons_dive_2024 %>% 
  ggplot(aes(x = meters_offshore, y = dive_time_obs)) + 
  geom_jitter() + 
  geom_smooth(method = "lm")
```

### Statistics
With both of our variables being continuous and numerical, we need to perform a regression test. First, we need to determine if there is a linear relationship between the variables. 

```{r}
loons_dive_2024 %>% 
  ggplot(aes(x = meters_offshore, y = dive_time_obs)) + 
  geom_jitter() + 
  geom_smooth(method = "lm") + 
  labs(title = "Dive Times vs. Meters Offshore", 
       subtitle = "of G. immer", 
       x = "Distance Offshore (m)", 
       y = "Dive Time (seconds)")
```

Next, we need to determine if both variables are normally distributed. 

```{r}
#Meters Offshore
ggplot(data = loons_dive_2024, mapping = aes(x = meters_offshore)) + 
  geom_histogram()

#Dive Times
ggplot(data = loons_dive_2024, mapping = aes(x = dive_time_obs)) + 
  geom_histogram()
```

#### Statistics
With the variable `meters_offshore` having a non-normal distribution, we will want to use a non-parametric version of a linear correlation test (i.e. Pearson's R). One such test is Spearman's Rho. 

```{r warning=FALSE}
dive_wave_corr <- cor.test(x=loons_dive_2024$meters_offshore, y = loons_dive_2024$dive_time_obs, method = 'spearman')

dive_wave_corr
```

REPORT: A Spearman'scorrelation indicates a moderate positive monotonic relationship between the variables `meters_offshore` and `dive_time_obs` (rs[93] = .50, p < 0.005). 





## Tidal height and dive times
The seventh analysis that will be conducted will be investigating the potential relationship between tidal height and *G. immer* dive times during the 2024 observation period. 

Hypothesis: Tidal height will significantly affect *G. immer* dive times.
Null Hypothesis: Tidal height will not significantly affect *G. immer* dive times.

### Visualizations
```{r}
loons_dive_2024 %>% 
  ggplot(aes(x = tidal_height, y = dive_time_obs)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

### Statistics
First, we need to see if our variables have a linear relationship

```{r}
loons_dive_2024 %>% 
  ggplot(aes(x = tidal_height, y = dive_time_obs)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

Next, we need to determine if both variables are normally distributed. 

```{r}
#tidal height
ggplot(data = loons_dive_2024, mapping = aes(x = tidal_height)) + 
  geom_histogram()

#dive times
ggplot(data = loons_dive_2024, mapping = aes(x = dive_time_obs)) + 
  geom_histogram()
```

#### Statistics
With the variable `tidal_height` having a non-normal distribution, we will want to use a non-parametric version of a linear correlation test, such as the Spearman's correlation. 

```{r warning=FALSE}
dive_height_corr <- cor.test(x=loons_dive_2024$tidal_height, y = loons_dive_2024$dive_time_obs, method = 'spearman')

dive_height_corr
```

REPORT: A Spearman's correlation indicates that there is no significant relationship between the variables `tidal_height` and `dive_time_obs` (rs[93] = -0.06, p = 0.54).

## Tidal classification and dive times

Hypothesis: Tidal class will significantly affect *G. immer* dive times.

Null Hypothesis: Tidal class will not significantly affect *G. immer* dive times.

### Visualizations

```{r}
loons_dive_2024$tide <- fct_relevel(loons_dive_2024$tide, c("mid_ebb",
                                                            "low_slack",
                                                            "low_flood",
                                                            "mid_flood",
                                                            "high_flood",
                                                            "high_slack",
                                                            "high_ebb"))
loons_dive_2024 %>% 
ggplot(mapping = aes(x = tide, y = dive_time_obs)) + 
  geom_boxplot()
```

### Statistics

```{r}
full_tide_aov = aov(dive_time_obs ~ tide, data = loons_dive_2024)
summary(full_tide_aov)
TukeyHSD(full_tide_aov)
```

An Analysis of Variance test demonstrated that tidal class has a significant effect on the length of *G. immer* dive times (ANOVA, F_6,86 = 4.599; p = 0.000427). An uplanned analytical comparison demonstrated that *G. immer* dive times at low slack tide were signficantly different than those at mid ebb tide (TukeyB, p < 0.0005055), and dive lengths at high ebb tide were significantly different than those at mid ebb tide (TukeyB, p = 0.0017457). Otherwise, there were no significant differences in dive times between other tidal classes (TukeyB, p > 0.05)

# Discussion / Conclusion







# Citations
Bent, A. C. 2009. Life Histories of North American Diving Birds. Cornell University Press: Ithaca, New York.

Daub, B. C. 1989. Behavior of Common Loons in Winter. Journal of Field Ornithology 60: 305-311.

Ford, T. B. and Gieg, J. A. 1995. Winter Behavior of the Common Loon. Journal of Field Ornithology 66: 22-29.

McIntyre, J. W. 1978. Wintering Behavior of Common Loons. The Auk 95: 396-403.

Holm, K. and Burger, A. 2002. Foraging Behavior and Resource Partitioning by Diving Birds During Winter in Areas of Strong Tidal Currents. Waterbirds 25(3): 312-325.

Thompson, Stephanie A. & Price, J. J. (2006). Water Clarity and Diving Behavior in Wintering Common Loons. Waterbirds: The International Journal of Waterbird Biology, 29(2), 169–175.
