---
title: "Common Loon Analysis"
author: "Ellie Gabrielson, Will Draxler, Autumn Pauly"
output: github_document
---
# Background Information
## Introduction
Although Common Loons (Gavia immer) have been studied extensively during the breeding season not much is known of their behavior on their ocean wintering grounds, where Loons spend a significant part of their lives (Bent, 2009). The few studies that do look at winter behavior often yield conflicting results (McIntyre, 1978; Daub, 1989; Ford and Gieg, 1995). The possible use of different shoreline habitats for activities such as feeding, sleeping, or taking shelter is worthy of study. McIntyre (1978) found common loons were semi-social and formed loose rafts and that loon behavior varied with tides, weather, and time of day. Daub (1989) and later Ford and Gieg (1995) found refuting evidence suggesting loons were neither social nor territorial, and did not change their behavior much by tide. Holm and Burger (2002) found that the general pattern for loons was that they foraged during slack and main-flow water. Additionally, the possible difference in dive times as a result of varying tidal heights is worthy of study. Thompson and Price (2006) observed that dive times were longer during low tide in comparison to other tidal stages, though more studies should be conducted to assess the use of tidal stage and behavior. 

Our general hypothesis when analyzing this data is that environmental conditions affect the behavior of the G. immer. The null hypothesis to this statement is that environmental conditions do not affect the behavior of the G. immer.

## Methods
In 2023, observations were conducted at nine established observation sites along the coastline of Mount Desert Island in Hancock County (Maine). Observations occurred from January 2023 to March 2023 on a biweekly basis, where each waterfowl individual was recorded as well as their behaviors, referencing Daub’s (1989) list of behaviors, and coordinate location. Locational coordinates were obtained by estimating the individual’s rough location digitally (accurate to + or - 10 m). Each transect site was visited for roughly 5-10 minutes, where data collection followed initial observation. Utilizing a Kestrel 2500, weather data was recorded accurately to + or - 1°C of temperatures and within 3% of wind speed. Referencing Daub’s (1989) list of behaviors, each species’ behavior during the observation was noted. The weather and environmental data recorded included wind speed and direction, barometer, cloud cover, precipitation, humidity, temperature, tidal percentage, and wave class (Beaufort Wind Speed). Tidal data was gathered from NOAA’s digital tidal charts. This study was repeated in 2024 with the addition of observational variables including species dive times and tidal height. 

# Load Packages and Data
```{r load-packages, message = FALSE}
#install.packages("tidyverse")
library(tidyverse)
library(rmarkdown)
library(scales)
library(ggridges)
library(rstatix)
library(parsnip)
```

```{r read-in-data}
loons_2023 <- read_csv("loons_2023.csv")
loons_2024 <- read_csv("loons_2024.csv")
```


We should bootstrap all of the tidal heights so that we have equal observations... maybe not? 

```{r bootstrapping}
#boot_loon_2023 <- loons_2023_behavior %>% 
 # specify(response = behavior) %>% 
 # generate(reps = 15000, type = "bootstrap")
```

# Tidal Condition and Behavior for 2023 and 2024

Hypothesis: Tidal height has a significant effect on G. immer behavior

Null Hypothesis: Tidal height will not have a significant effect on G. immer behavior

```{r sampling-tides-behavior}
# #this is the dataset that we want to sample from? 
# loon_tide_behavior <- loons_2023 %>% 
#   select(behavior, tide)

## Tide-Behavior visualizations - to be moved to analysis
loons_2023 %>% 
  count(behavior, tide) %>% 
  ggplot(aes(x = behavior, y = n)) +
  geom_col() +
  facet_wrap(~ tide) + 
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ 
  labs(title = "Count of Behavior", 
       subtitle = "by Tidal Stage", 
       x = "Behavior", 
       y = "Count")

#tidal stage count graph
loons_2023 %>% 
  count(tide) %>% 
  ggplot(aes(x = tide, y = n)) +
      geom_col() + 
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(title = "Count of Observations", 
       subtitle = "by Tidal Stage", 
       x = "Tidal Stage", 
       y = "Count")
```

### Chi-Squared Test of Independence
The Chi-Squared Test of Independence examines whether there is a relationship between two categorical variables. Our null hypothesis, that tidal height will not have a significant effect on G. immer behavior, assumes that tidal condition and behavior are independent. Our hypothesis, that tidal height will have a significant effect on G. immer behavior, assumes that tidal condition and behavior are dependent.

#### Expected Frequencies and Contingency Table
Before performing the chi-square test, we need to calculate the expected frequencies for each cell in the contingency table. These expected frequencies represent what you would expect to observe if there were no association between tidal condition and behavior.

```{r contingency-table}
contingency_table1 <- table(loons_2023$tide, loons_2023$behavior)
contingency_table1
```

#### Chi-square Test
Now that we have the contingency table of expected frequencies, we can perform the statistical test. 

```{r chi-square-test}
chi_sq_test1 <- chisq.test(contingency_table1)
chi_sq_test1

# # x-squared
# chi_sq_test1$statistic
# 
# #p-value
# chi_sq_test1$p.value
# 
# #expected frequencies
# chi_sq_test1$expected
```

REPORT: The behavior of Common Loons is dependent on tidal condition (Pearson's Chi-squared test; X-squared = 33.184, df = 10, p-value > 0.005). 

```{r}
install.packages("summarytools")
library(summarytools)
library(dplyr)

# fourth method:
loons_2023 %$%
  ctable(tide, behavior,
    prop = "r", chisq = TRUE, headings = FALSE
  ) %>%
  print(
    method = "render",
    style = "rmarkdown",
    footnote = NA
  )
```

<!-- #### ANOVA -->
<!-- If we want to compare the abundance of diving behavior between the three tidal stages (high tide, mid tide, and low tide), we need to compare the means of a count variable (number of diving behaviors) across different groups (tidal stages). ANOVA is suitable for comparing means across multiple groups, which makes it appropriate for this scenario. -->

<!-- ```{r} -->
<!-- loon_2023_diving <- loons_2023 %>% -->
<!--   select(species, behavior, date, tide, overall_abundance, specific_abundance) %>% -->
<!--   filter(species == "COMMON LOON") %>% -->
<!--   mutate(behavior = if_else(behavior == "DIVING", "diving", "other")) %>% -->
<!--   group_by(date, overall_abundance, tide) %>% -->
<!--   count(behavior) %>% -->
<!--   mutate(diving_count = n) -->

<!-- loon_2023_diving <-loon_2023_diving %>% -->
<!--   filter(behavior == "diving") %>% -->
<!--   mutate(diving_percentage = diving_count/overall_abundance) -->
<!-- ``` -->

<!-- Running an ANOVA test -->

<!-- ```{r} -->
<!-- # Perform ANOVA -->
<!-- anova_model1 <- aov(diving_percentage ~ tide, data = loon_2023_diving) -->
<!-- summary(anova_model1) -->

<!-- # Perform Tukey's HSD test for pairwise comparisons -->
<!-- tukey_results1 <- TukeyHSD(anova_model1) -->
<!-- print(tukey_results1) -->
<!-- ``` -->

<!-- REPORT: -->

# Tidal Conditions and Dive Times - 2024

Hypothesis: Low tide will cause G. immer to have significantly longer dive times. 

Null Hypothesis: Low tide will not cause G. immer to have significantly longer dive times. 

```{r mutation, baby}
library(stringr)

loons_dive_2024 <- loons_2024 %>% 
  filter(species == "common_loon") %>% 
  select(tide, behavior, dive_time_obs, tidal_height) %>% 
  filter(dive_time_obs != "n/a") %>% 
  filter(dive_time_obs != "na")

loons_dive_2024 <- loons_dive_2024 %>% 
  separate_longer_delim(dive_time_obs, delim = ",")

loons_dive_2024 <- loons_dive_2024 %>%
  mutate(dive_time_obs = as.numeric(dive_time_obs))

loons_dive_2024 <- loons_dive_2024 %>% 
  mutate(general_tide = fct_recode(tide, "low" = "low_flood", 
                           "low" = "low_slack",
                           "mid" = "mid_flood",
                           "mid" = "mid_ebb",
                           "high" = "high_flood",
                           "high" = "high_slack",
                           "high" = "high_ebb"))
```

```{r}
loons_dive_2024$tide.f <- fct_relevel(loons_dive_2024$general_tide, c("low", "mid", "high"))

ggplot(data = loons_dive_2024, mapping = aes(x = tide.f, y = dive_time_obs)) + 
  geom_boxplot() +
  labs(title = "Tidal Conditions and Dive Times", 
       subtitle = "of Gavia immer", 
       x = "Tidal Stage", 
       y = "Dive Time (seconds)")
```

```{r}
woomaov = aov(dive_time_obs ~ general_tide, data = loons_dive_2024)
summary(woomaov)
TukeyHSD(woomaov)

#high to low
loons_dive_2024_1 <- loons_dive_2024 %>% 
  filter(general_tide == "high"|general_tide == "mid")

t.test(dive_time_obs ~ general_tide, data = loons_dive_2024_1)

#low to high
loons_dive_2024_1 <- loons_dive_2024 %>% 
  filter(general_tide == "high"|general_tide == "low")

t.test(dive_time_obs ~ general_tide, data = loons_dive_2024_1)

#low to mid
loons_dive_2024_1 <- loons_dive_2024 %>% 
  filter(general_tide == "mid"|general_tide == "low")

t.test(dive_time_obs ~ general_tide, data = loons_dive_2024_1)
```
REPORT: There was no significant difference between dive times during low tides and high tides [insert test stat] and between dive times during mid tides and high tides [insert test statistic]. The dive times during mid tides were significantly lower than dive times during low tides [insert test statistic]. 

The average dive time during high tide was 42.02 seconds, 30.6 seconds during mid tides, and 47.4 seconds during low tides. 

```{r adjusted-r-squared}

#setting linear model- using tidal height
times_vs_tide_model <- linear_reg() %>% 
  set_engine("lm") %>% 
  fit(dive_time_obs ~ tidal_height, data = loons_dive_2024)

#augmenting the weight data
times_vs_tide_model_augment <- augment(times_vs_tide_model$fit)

ggplot(times_vs_tide_model_augment, mapping = aes(x = .fitted, y = .resid)) +
  geom_jitter(alpha = 0.75) +
  geom_smooth(color = "black") +
  labs(x = "Predicted Dive Time", y = "Residuals")

#assessing the r-squared value
glance(times_vs_tide_model)$r.squared
```
R_SQUARED REPORT: The r-squared value is `[...]`, which means that the variable `general_tide` accounts for [...] of the variation in the dependent variable, `dive_time_obs`. 

# Exposure Level and Abundance - 2023

Hypothesis: Higher exposure level of location significantly positively affects G. immer presence. 

Null Hypothesis: Higher exposure level of location will not significantly positively affect G. immer presence. 

Predictor variable = categorical
outcome variable = quantitative
test = ANOVA

```{r}
loons_2023 %>% 
  ggplot(mapping = aes(x = shelter_gradient, y = specific_abundance)) + 
  geom_boxplot() +
  labs(title = "Common Loon Abundance", 
       subtitle = "by Shelter Gradient", 
       x = "Shelter Gradient", 
       y = "Abundance")
```

```{r}
loons_2023 %>% 
  ggplot(mapping = aes(x = specific_abundance)) +
  geom_histogram() +
  geom_density() + 
  facet_wrap(~shelter_gradient, ncol = 2)

#plot
loons_2023 %>% 
  ggplot(aes(x = specific_abundance, y = shelter_gradient, fill = shelter_gradient)) + 
  geom_density_ridges(alpha = 0.6, bins = 1) + 
  theme_ridges() + 
   theme(legend.position = "none", panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8))+
  theme_minimal() +
  scale_fill_viridis_d() +
  labs(title = "Common Loon Abundance", 
       subtitle = "by Shelter Gradient", 
       x = "Abundance", 
       y = "Shelter Gradient", 
       fill = "Shelter Gradient")
```

```{r}
abundanceexposureaov = aov(specific_abundance ~ shelter_gradient, data = loons_2023)
summary(abundanceexposureaov)
TukeyHSD(abundanceexposureaov)
```

REPORT: An Analysis of Variance test demonstrated that coastal exposure levl has a significant effect on the abundance of G. immer (ANOVA, F_4,126 = 6.167; p = 0.000145). 
An uplanned analytical comparison demonstrated that G. immer prescence

# Exposure Level and Behavior

Hypothesis: Exposure level of location significantly impacts G. immer behavior. 

Null Hypothesis: Exposure level of location has no significant impact on G. immer behavior. 


```{r}
loons_2023 %>% 
  count(behavior, shelter_gradient) %>% 
  ggplot(aes(x = behavior, y = n)) +
  geom_col(fill = "firebrick4", color = "black") +
  facet_wrap(~ shelter_gradient, nrow = 1) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Count of Common Loon Behavior", 
       subtitle = "by Shelter Gradient", 
       x = "Behavior", 
       y = "Count")
```

```{r}
behavior_exposure_kruskal <- kruskal.test(behavior ~ shelter_gradient, data = loons_2023)
behavior_exposure_kruskal

behavior_exposure_dunn <- loons_2023 %>% dunn_test(behavior ~ shelter_gradient, p.adjust.method = "bonferroni")
behavior_exposure_dunn
```

### REPORT

#Wave Class and Behavior

Hypothesis: Wave class will significantly affect G. immer behavior.

Null Hypothesis: Wave class will not significantly affect G. immer behavior.

```{r}
loons_2023 %>% 
 ggplot(aes(x = wave_class)) +
  geom_histogram() +
  facet_wrap(~ behavior, nrow = 1)

loons_2023 %>%
  ggplot(aes(x = wave_class, y = behavior, fill = behavior)) + 
  geom_density_ridges(alpha = 0.6, bins = 1)

waveclassaov = aov(wave_class ~ behavior, data = loons_2023)
summary(waveclassaov)
  
```
Wave class did not significantly affect G. immer behavior (one-way ANOVA, F_4,126 = 2.356, P = 0.0572).

#Meters Offshore and Dive Times

Hypothesis: Distance from shore will significantly affect G. immer dive times. 

Null Hypothesis: Distance from shore will not significantly affect G. immer dive times.

```{r}
loons_dive_2024 <- loons_2024 %>% 
  filter(species == "common_loon") %>% 
  select(tide, behavior, dive_time_obs, tidal_height, meters_offshore) %>% 
  filter(dive_time_obs != "n/a") %>% 
  filter(dive_time_obs != "na")

loons_dive_2024 <- loons_dive_2024 %>% 
  separate_longer_delim(dive_time_obs, delim = ",") %>% 
  mutate(dive_time_obs = as.numeric(dive_time_obs))

loons_dive_2024 %>% 
  ggplot(aes(x = meters_offshore, y = dive_time_obs)) + 
  geom_jitter() + 
  geom_smooth(method = "lm")

#linear model
loon.dive.meters.lm <- lm(dive_time_obs ~ meters_offshore, data = loons_dive_2024)

summary(loon.dive.meters.lm)

```

A linear model demonstrated a significant positive relationship between dive time and meters off shore (F1,89 = 27.81 : p = <0.5). 24% of the total variability in dive time can be predicted through a knowledge of meters off shore.

#Tidal height and dive times

Hypothesis: Tidal height will significantly affect G. immer dive times.

Null Hypothesis: Tidal height will not significantly affect G. immer dive times.

```{r}
loons_dive_2024 %>% 
  ggplot(aes(x = tidal_height, y = dive_time_obs)) + 
  geom_point() +
  geom_smooth(method = "lm")


dive_time_tide_height_lm <- lm(dive_time_obs ~ tidal_height, data = loons_dive_2024)
summary(dive_time_tide_height_lm)

#This is actually so interesting
```

A linear model demonstrated no significant relationship between dive time and tidal height (F1,88 = 0.014 : p = 0.91).

#Tidal classification and dive times

Hypothesis: Tidal class will significantly affect G. immer dive times.

Null Hypothesis: Tidal class will not significantly affect G. immer dive times.

```{r}
loons_dive_2024$tide <- fct_relevel(loons_dive_2024$tide, c("mid_ebb",
                                                            "low_slack",
                                                            "low_flood",
                                                            "mid_flood",
                                                            "high_flood",
                                                            "high_slack",
                                                            "high_ebb"))
loons_dive_2024 %>% 
ggplot(mapping = aes(x = tide, y = dive_time_obs)) + 
  geom_boxplot()
full_tide_aov = aov(dive_time_obs ~ tide, data = loons_dive_2024)
summary(full_tide_aov)
TukeyHSD(full_tide_aov)
```

An Analysis of Variance test demonstrated that tidal class has a significant effect on the length of G. immer dive times (ANOVA, F_6,86 = 4.599; p = 0.000427). An uplanned analytical comparison demonstrated that G. immer dive times at low slack tide were signficantly different than those at mid ebb tide (TukeyB, p < 0.0005055), and dive lengths at high ebb tide were significantly different than those at mid ebb tide (TukeyB, p = 0.0017457). Otherwise, there were no significant differences in dive times between other tidal classes (TukeyB, p > 0.05)

# Discussion / Conclusion

# Citations
Bent, A. C. 2009. Life Histories of North American Diving Birds. Cornell University Press: Ithaca, New York.

Daub, B. C. 1989. Behavior of Common Loons in Winter. Journal of Field Ornithology 60: 305-311.

Ford, T. B. and Gieg, J. A. 1995. Winter Behavior of the Common Loon. Journal of Field Ornithology 66: 22-29.

McIntyre, J. W. 1978. Wintering Behavior of Common Loons. The Auk 95: 396-403.

Holm, K. and Burger, A. 2002. Foraging Behavior and Resource Partitioning by Diving Birds During Winter in Areas of Strong Tidal Currents. Waterbirds 25(3): 312-325.

Thompson, Stephanie A. & Price, J. J. (2006). Water Clarity and Diving Behavior in Wintering Common Loons. Waterbirds: The International Journal of Waterbird Biology, 29(2), 169–175.
