---
title: "Common Loon Data Tidying"
subtitle: "for Statistics Final Project"
author: "Ellie Gabrielson, Will Draxler, Autumn Pauly"
date: "03/12/2024"
output: 
  html_document: 
    toc: true
---
## Loading the Packages and Data

We will be using the `tidyverse` and `lubridate` packages to assist with data tidying. 

```{r loading-packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
```

## Loading Data

First we need to load in the data that was collected in 2023 and 2024. 

```{r loading-data, warning=FALSE, message=FALSE}
#2023
all_bird_untidy_2023 <- read_csv("loons_2023.csv")
loons_2023 <- read_csv("loons_2023.csv")


```

## Tidying the Data
### Exposure

The nine locations that are going to be used to assess the environmental preferences of the Common Loon need to be refactored by exposure level. 

```{r refactor-exposure-level, message=FALSE, warning=FALSE}
#assigning exposure level to locations
loons_2023 <- loons_2023 %>% 
  mutate(shelter_gradient = case_when(location == "BAR HARBOR PIER" ~ "exposed",
                                      location == "SEAL HARBOR BEACH" ~ "moderate",
                                      location == "BRACY HARBOR" ~ "moderate",
                                      location == "NORTHEAST HARBOR" ~ "moderately_sheltered",
                                      location == "SOMES SOUND" ~ "sheltered",
                                      location == "SOUTHWEST HARBOR" ~ "moderately_sheltered",
                                      location == "SEAWALL" ~ "exposed",
                                      location == "SEAL COVE" ~ "moderately_exposed",
                                      location == "SAND BEACH" ~ "moderately_exposed"))
 
#releveling the exposure levels from most exposed to least exposed 
shelter_factor <- fct_relevel(loons_2023$shelter_gradient, c(
                "exposed", 
                "moderately_exposed",
                "moderate",
                "moderately_sheltered",
                "sheltered"))
```

After tidying this, we need to save these changes into our dataset. 

```{r saving-new-dataset, message=FALSE, warning=FALSE}
#saving a dataset to include only loons and the shelter factor
loons_2023 <- loons_2023 %>% 
  mutate(shelter_factor = shelter_factor) %>% 
  filter(species == "COMMON LOON")

loons_2023 <- loons_2023 %>% 
  mutate(general_tide = case_when(tide == "LOW" ~ "low",
                                      tide == "MID" ~ "mid",
                                      tide == "HIGH" ~ "high"))

loons_2023 <- loons_2023 %>% 
  mutate(behavior = case_when(behavior == "PEERING" ~ "peering",
                              behavior == "DIVING" ~ "diving",
                              behavior == "DRIFTING" ~ "drifting",
                              behavior == "MAINTENANCE" ~ "maintenance", 
                              behavior == "ASHORE" ~ "ashore"
                                      ))
```



### Changing `chr` variables to `num`

There are a number of variables within this dataset that are character variables that should be numeric variables (i.e. meters_offshore, longitude, latitude)

```{r chr-to-num, warning=FALSE, message=FALSE}
#Change character numbers to numeric
loons_2023 <- loons_2023 %>% 
  mutate(meters_offshore = as.numeric(meters_offshore)) %>% 
  mutate(latitude = as.numeric(latitude)) %>% 
  mutate(longitude = as.numeric(longitude)) %>% 
  mutate(barometer = as.numeric(barometer)) %>% 
  mutate(date = as_date(date, format = "%m/%d/%Y")) %>% 
  mutate(time = hms::as_hms(time))
```


###Tidy gis based data
```{r}
# dives_gis <- read_csv("loons_dive_2024_gis_temp.csv") 
#   
# dives_gis1 <- dives_gis %>% 
#   filter(is.na(`RASTERVALU...10`) == F ) %>% 
#   mutate(bathy_depth_m = `RASTERVALU...10`) %>% 
#   select(-`RASTERVALU...10`)
# 
# dives_gis2 <- dives_gis %>% 
#   filter(is.na(`RASTERVALU...9`) == F ) %>% 
#   mutate( bathy_depth_m = `RASTERVALU...9`) %>% 
#   select(-`RASTERVALU...9`)
# 
# dives_gis1 %>% 
#   bind_rows(dives_gis2) %>% 
#   select(-RASTERVALU) %>%
#   mutate(water_depth_m = (bathy_depth_m * -1) + tidal_height) %>% 
#   write_csv("dives_gis_tidy.csv")
```


# Writing CSV

Now, we should write this new tidied dataset as a CSV so that it can be read across all the .Rmd files. 

```{r CSV-writing}
write_csv(loons_2023, "loon_2023_tidy.csv")

```